---
title: "50-1183847669 CoCl2-treated THP-1 Data Analysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---


# Gene count analysis with DESEQ2 Control vs CoCl2-treated THP-1

## Set up enviroment

```{r}
library(tidyverse)
library(DESeq2)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(pheatmap)
library(stringr)
library(ReactomePA)
library(matrixStats)
library(msigdbr)
library(fgsea)
library(ggplot2)
library(dplyr)
library(writexl)
library(tibble)
library(readr)
```


# 1. Load counts matrix and gene symbols

```{r}
counts <- read_tsv("/path_to_input/featurecounts_counts_with_symbols.tsv")
counts_df <- as.data.frame(counts)
head(counts_df)
```


```{r}
# Removing first five rows as they are not required for analysis
counts_df <- counts_df[,-(2:6) ]
head(counts_df)
```

```{r}
# rename columns 
colnames(counts_df) <- c("Geneid", "Control 1", "Control 2", "Control 3", "TreatA1", "TreatA2", "TreatA3", "CoCl2 1", "CoCl2 2", "CoCl2 3", "gene_symbol")
```


```{r}
# add rownames and keep only samples for comparing control to treated
rownames(counts_df) <- counts_df$Geneid
counts_matrix <- counts_df %>% dplyr::select(-c(Geneid, gene_symbol))
counts_matrix <- counts_matrix[, !colnames(counts_matrix) %in% c("TreatA1", "TreatA2", "TreatA3")]
head(counts_matrix)
```


# 2. Inspect total gene counts per sample


```{r}
total_counts <- colSums(counts_matrix %>% dplyr::select(where(is.numeric)))
total_counts_df <- tibble(Sample = names(total_counts), TotalCounts = total_counts)
ggplot(total_counts_df, aes(x = Sample, y = TotalCounts, fill = Sample)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title="Raw Gene Counts", y="Total Counts", x="Sample") +
  scale_fill_brewer(palette="Set1") +
  guides(fill=FALSE)
```


# 3. Create metadata and DESeq2 object

```{r}
# Filter lowly expressed genes and keep genes with counts >= 10 in at least 2 samples
keep <- rowSums(counts_matrix >= 10) >= 2
counts_matrix <- counts_matrix[keep, ]
dim(counts_matrix)
```

26,744 genes


```{r}
# create dds object
colData <- data.frame(
  row.names = colnames(counts_matrix),
  condition = c("control","control","control","treated","treated","treated") 
)

dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = colData,
                              design = ~ condition)
```


```{r}
# Run DESeq2
dds <- DESeq(dds)

# normalize the counts
norm_counts <- counts(dds, normalized=TRUE)
```

```{r}
normalized_counts <- as.data.frame(norm_counts)
head(normalized_counts)
```
```{r}
raw_counts <- as.data.frame(counts_matrix)
head(raw_counts)
```


```{r}
# export raw counts
raw_counts <- raw_counts %>%
  rownames_to_column(var = "GeneID")

colnames(raw_counts) <- gsub(" ", "_", colnames(raw_counts))

write_csv(raw_counts,
          "/path_to_output/50-1183847669_CoCl2_THP1_TREAT_VS_CONTROL_raw_gene_counts.csv")
```


```{r}
# export normalized counts
normalized_counts <- normalized_counts %>%
  rownames_to_column(var = "GeneID")

colnames(normalized_counts) <- gsub(" ", "_", colnames(normalized_counts))

write_csv(normalized_counts,
          "/path_to_output/50-1183847669_CoCl2_THP1_TREAT_VS_CONTROL_normalized_gene_counts.csv")
```



```{r}
# Visualise normalized counts
total_norm_counts <- colSums(normalized_counts %>% dplyr::select(where(is.numeric)))
total_norm_counts_df <- tibble(
  Sample = names(total_norm_counts),
  TotalCounts = total_norm_counts
)
ggplot(total_norm_counts_df, aes(x = Sample, y = TotalCounts, fill = Sample)) +
  geom_bar(stat = "identity", color = "black") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Normalized Gene Counts", y = "Total Normalized Counts", x = "Sample") +
  scale_fill_brewer(palette = "Set1") +
  guides(fill = FALSE)
```



```{r}
df_long <- as.data.frame(norm_counts) %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to="sample", values_to="normalized_count")
ggplot(df_long, aes(x=sample, y=normalized_count)) +
  geom_boxplot(fill="lightgreen") +
  theme_minimal() +
  xlab("Sample") +
  ylab("Normalized Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




# 4. Principal Component Anlysis

```{r}
# Run PCA
vsd <- vst(dds, blind=FALSE)
plotPCA(vsd, intgroup="condition")
```
```{r}
# Extract PCA data
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color = condition)) +
  geom_point(size = 4, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal(base_size = 16) +
  theme(
    text = element_text(family = "Times New Roman"),  
    legend.position = "right",
    panel.grid.major = element_line(color = "gray95"),
    panel.grid.minor = element_blank()
  ) +
  coord_fixed(ratio = 1/0.35)  
```
## Top Variable Genes

```{r}
# vsd is a DESeqTransform object
vsd_mat <- assay(vsd) 
topVarGenes <- head(order(rowVars(vsd_mat), decreasing = TRUE), 50)
vsd_top <- vsd_mat[topVarGenes, ]
pheatmap(vsd_top, 
         scale = "row",          
         annotation_col = as.data.frame(colData(vsd)),
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE)

```

## Add Gene Symbol to replace "EN..." with actual gene name

```{r}
gene_symbols <- counts_df$gene_symbol[match(rownames(vsd_mat), counts_df$Geneid)]
rownames(vsd_mat) <- gene_symbols
topVarGenes <- head(order(rowVars(vsd_mat), decreasing = TRUE), 65)
vsd_top <- vsd_mat[topVarGenes, ]
```


```{r}
# Plot heatmap
pheatmap(vsd_top, 
         scale = "row",
         annotation_col = as.data.frame(colData(vsd)),
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         fontsize = 12,         
         family = "Times New Roman")  

```



# 5. Differential Gene Expression


```{r}
res <- results(dds, alpha = 0.05) 
res <- res[!is.na(res$padj), ] # removing any NAs
res$gene_symbol <- counts_df$gene_symbol[match(rownames(res), counts_df$gene_id)] # adding gene symbol 
```

```{r}
# Shrink log2FC 
resLFC <- lfcShrink(dds, coef=2, type="apeglm") 
resLFC$gene_symbol <- counts_df$gene_symbol[match(rownames(resLFC), counts_df$gene_id)]
```


```{r}
# Strip version numbers from rownames + Geneid
res$Geneid_clean <- sub("\\.\\d+$", "", rownames(res))
resLFC$Geneid_clean <- sub("\\.\\d+$", "", rownames(resLFC))
counts_df$Geneid_clean <- sub("\\.\\d+$", "", counts_df$Geneid)

res$gene_symbol <- counts_df$gene_symbol[match(res$Geneid_clean, counts_df$Geneid_clean)]
resLFC$gene_symbol <- counts_df$gene_symbol[match(resLFC$Geneid_clean, counts_df$Geneid_clean)]
sum(!is.na(resLFC$gene_symbol)) # check if worked
```

```{r}
head(res)
```
```{r}
head(resLFC)
```



# 6. Volcano plot with EnhancedVolcano


```{r}
library(EnhancedVolcano)

EnhancedVolcano(resLFC,
    lab = rep("", nrow(resLFC)),  # give empty strings instead of NULL
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    pointSize = 3.0,
    labSize = 0,         # ensures no visible labels
    colAlpha = 0.7,
    legendLabels = c('NS','Log2FC','padj','padj & Log2FC'),
    legendPosition = 'right',
    drawConnectors = FALSE,  # no lines since no labels
    title = 'Control vs. CoCl2-treated',
    subtitle = 'Volcano plot of differentially expressed genes',
    xlim = c(-10, 10),
    caption = NULL) +
  theme(
    text = element_text(family = "Times New Roman")  # Times New Roman
  )

```

```{r}
# Getting the number of significantly up and down regulated genes with LFC > 1.5
padj_cutoff <- 0.05
lfc_cutoff  <- 1.5

# Upregulated genes
up_count <- sum(resLFC$padj < padj_cutoff & resLFC$log2FoldChange >  lfc_cutoff, na.rm = TRUE)
cat("Upregulated genes:", up_count, "\n")

# Downregulated genes
down_count <- sum(resLFC$padj < padj_cutoff & resLFC$log2FoldChange < -lfc_cutoff, na.rm = TRUE)
cat("Downregulated genes:", down_count, "\n")
```



```{r}
# labeling top up and down regulated genes
top_up <- resLFC %>%
  as.data.frame() %>%
  filter(!is.na(gene_symbol) & log2FoldChange > 0) %>%
  arrange(padj) %>%
  head(500) %>% # can change the number here depending on how many genes you want to visualize
  pull(gene_symbol)

top_down <- resLFC %>%
  as.data.frame() %>%
  filter(!is.na(gene_symbol) & log2FoldChange < 0) %>%
  arrange(padj) %>%
  head(500) %>%
  pull(gene_symbol)

genes_to_label <- c(top_up, top_down)

EnhancedVolcano(resLFC,
                lab = resLFC$gene_symbol,
                x = 'log2FoldChange',
                y = 'padj',
                selectLab = genes_to_label,  
                pCutoff = 0.05,
                FCcutoff = 1.5,
                pointSize = 3.0,
                labSize = 4.0,
                colAlpha = 0.7,
                legendLabels=c('NS','Log2FC','padj','padj & Log2FC'),
                legendPosition = 'right',
                drawConnectors = TRUE,
                widthConnectors = 0.5,
                colConnectors = 'grey30',
                title = 'Treated vs Control',
                subtitle = 'Volcano plot of DE genes',
                xlim = c(-10, 10))
```



# 7. Gene Ontology enrichment with clusterProfiler

```{r}
# Subset all the significant genes (up and downregulated) (adjusted p-value < 0.05)
sig_genes <- unique(resLFC$gene_symbol[!is.na(resLFC$padj) & resLFC$padj < 0.05])

# Map the subsetted genes to Entrez IDs
gene_df <- bitr(sig_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
entrez_ids <- gene_df$ENTREZID

# GO enrichment analysis (Biological Process)
ego <- enrichGO(gene         = entrez_ids,
                OrgDb        = org.Hs.eg.db,
                ont          = "BP",
                pAdjustMethod = "BH",
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego_df <- as.data.frame(ego)

# Sort by adjusted p-value first, then by Count (descending)
ego_top <- ego_df %>%
  arrange(p.adjust, desc(Count)) %>%
  head(20)  # can change if you want to see more or less terms

# Dotplot of results
ggplot(ego_top, aes(x = reorder(Description, -Count), y = Count, 
                    color = -log10(p.adjust), size = Count)) +
  geom_point() +
  coord_flip() +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Top GO Biological Processes",
       x = "GO Term",
       y = "Gene Count",
       color = "-log10(adj. p-value)",
       size = "Gene Count") +                                                                 
  theme_minimal(base_size = 14)
```


```{r}
# Dotplot visualisation
ego_top$Description <- str_wrap(ego_top$Description, width = 120)
ggplot(ego_top, aes(x = reorder(Description, Count), y = Count, 
                    color = -log10(p.adjust), size = Count)) +
  geom_point() +
  coord_flip() +
  scale_color_gradient(low = "palegreen", high = "darkgreen") +
  labs(
    title = "Top GO Biological Processes",
    x = "",
    y = "Gene Count",
    color = "-log10(adj. p-value)",
    size = "Gene Count"
  ) +
  theme_minimal(base_size = 14, base_family = "Times New Roman") +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 22, face = "bold", color = "black"),
    axis.text.x  = element_text(size = 15, color = "black"),
    axis.text.y  = element_text(size = 13, color = "black", hjust = 1),  # adjust horizontal justification
    axis.title.x = element_text(size = 13, color = "black", face = "bold"),
    axis.title.y = element_text(size = 13, color = "black", face = "bold"),
    legend.title = element_text(size = 15, color = "black"),
    legend.text  = element_text(size = 12, color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.5)
  )
```

```{r}
# Boxplot visualisation
ego_top$Description <- str_wrap(ego_top$Description, width = 100)  

ggplot(ego_top, aes(x = reorder(Description, Count), y = Count)) +
  geom_col(fill = "darkgreen", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Top GO Biological Processes",
    x = "",
    y = "Gene Count"
  ) +
  theme_minimal(base_size = 14, base_family = "Times New Roman") +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 22, face = "bold", color = "black"),
    axis.text.x  = element_text(size = 15, color = "black"),
    axis.text.y  = element_text(size = 13, color = "black", hjust = 1),  # left-align long labels
    axis.title.x = element_text(size = 13, color = "black", face = "bold"),
    axis.title.y = element_text(size = 13, color = "black", face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.5)
  )

```




## Splitting by up or downregulated genes

```{r}
up_genes <- resLFC$gene_symbol[resLFC$padj < 0.05 & resLFC$log2FoldChange > 0]
down_genes <- resLFC$gene_symbol[resLFC$padj < 0.05 & resLFC$log2FoldChange < 0]

# Map to Entrez IDs
up_entrez <- bitr(up_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID
down_entrez <- bitr(down_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID

# run analysis on up and down genes
ego_up <- enrichGO(
  gene = up_entrez,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)

ego_down <- enrichGO(
  gene = down_entrez,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)
```



```{r}
plot_go_direction <- function(enrich_result, title, color = "red", n = 20, wrap_width = 100) {
  
  df <- as.data.frame(enrich_result) %>%
    filter(Count > 50) %>%          
    arrange(p.adjust, desc(Count)) %>%
    head(n)
  
  df$Description <- stringr::str_wrap(df$Description, width = wrap_width)
  
  ggplot(df, aes(x = reorder(Description, Count), y = Count)) +
    geom_point(aes(color = -log10(p.adjust), size = Count), alpha = 0.8) +
    coord_flip() +
    scale_color_gradient(low = "lightgray", high = color) +
    labs(
      title = title,
      x = "",
      y = "Gene Count",
      color = "-log10(adj. p-value)",
      size = "Gene Count"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      text = element_text(family = "Times New Roman"),
      axis.title = element_text(face = "bold", size = 14, color = "black"),
      axis.text.y = element_text(size = 12, color = "black", hjust = 1),  # moderate wrapping
      axis.text.x = element_text(size = 12, color = "black"),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "black"),
      legend.title = element_text(face = "bold", color = "black"),
      legend.text = element_text(color = "black"),
      panel.grid = element_blank(),
      axis.line = element_line(color = "black", size = 0.5)
    )
}
```

```{r}
plot_go_direction(ego_up, title = "Top GO BP (Upregulated)", color = "red")
plot_go_direction(ego_down, title = "Top GO BP (Downregulated)", color = "blue")
```


# KEGG enrichment


```{r}
kegg_up   <- enrichKEGG(gene = up_ids, organism = "hsa")
kegg_down <- enrichKEGG(gene = down_ids, organism = "hsa")

```

```{r}
plot_top_terms <- function(enrich_result, title, n=20) {
  df <- as.data.frame(enrich_result) %>%
    arrange(p.adjust, desc(Count)) %>%
    head(n)
  
  ggplot(df, aes(x = reorder(Description, Count), 
                 y = Count,
                 color = -log10(p.adjust),
                 size = Count)) +
    geom_point(alpha = 0.9) +
    coord_flip() +
    scale_color_gradient(low = "blue", high = "red") +
    labs(title = title,
         x = "GO Term",
         y = "Gene Count",
         color = expression(-log[10]~"(adj. p-value)"),
         size = "Gene Count") +
    theme_minimal(base_size = 14, base_family = "Times New Roman") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 11),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
}
```

```{r}
plot_top_terms(kegg_up,   "Upregulated Genes - KEGG Pathways")
plot_top_terms(kegg_down, "Downregulated Genes - KEGG Pathways")
```
```{r}
# 5. Reactome Pathway Enrichment
react_up   <- enrichPathway(gene = up_ids, organism = "human", readable=TRUE)
react_down <- enrichPathway(gene = down_ids, organism = "human", readable=TRUE)
```


```{r}
plot_top_terms(react_up,   "Upregulated Genes - Reactome Pathways")
plot_top_terms(react_down, "Downregulated Genes - Reactome Pathways")
```



# 9. Gene enrichment analysis with GSEA

```{r}
library(fgsea)
library(msigdbr)

# prepare DESeqResults
res_df <- as.data.frame(resLFC)
res_df <- res_df[!is.na(res_df$gene_symbol) & is.finite(res_df$log2FoldChange), ] # Keep only rows with valid gene symbols and finite log2FoldChange
res_df <- res_df[!duplicated(res_df$gene_symbol), ] # Keep only one entry per gene symbol

# Make a ranked gene vector
ranked_genes <- res_df$log2FoldChange
names(ranked_genes) <- res_df$gene_symbol
stopifnot(all(is.finite(ranked_genes)))

# Load Hallmark gene sets
m_df <- msigdbr(species = "Homo sapiens", category = "H")
pathways <- split(x = m_df$gene_symbol, f = m_df$gs_name)

# Run FGSEA
fgseaRes <- fgsea(
  pathways = pathways,
  stats = ranked_genes,
  minSize = 15,
  maxSize = 500#,
  #nperm = 10000
)

# Top pathways
topPathways <- fgseaRes %>% arrange(padj) %>% head(20)
topPathways

```

```{r}
ggplot(topPathways, aes(x = reorder(pathway, NES), y = NES, fill = NES)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Top Enriched Hallmark Pathways",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)",
    fill = "NES"
  ) +
  theme_minimal(base_size = 14, base_family = "Times New Roman") +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x  = element_text(size = 12),
    axis.text.y  = element_text(size = 12),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 11)
  )
```



```{r}
# Filter pathways to highlight the key pathways we are interested in related to hypoxia, immune response, dna damage, mitochondria and metabolism
exclude <- c("HALLMARK_MYOGENESIS", "HALLMARK_ADIPOGENSIS", "HALLMARK_ESTROGEN_RESPONSE_LATE", "HALLMARK_HEME_METABOLISM", "HALLMARK_XENOBIOTIC_METABOLISM",
             "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_ADIPOGENESIS", "HALLMARK_ESTROGEN_RESPONSE_EARLY", "HALLMARK_MYC_TARGETS_V2", "HALLMARK_MYC_TARGETS_V1",
             "HALLMARK_E2F_TARGETS", "HALLMARK_APICAL_JUNCTION", "HALLMARK_FATTY_ACID_METABOLISM")

topPathways_filtered <- topPathways %>%
  filter(!pathway %in% exclude)

ggplot(topPathways_filtered, aes(x = reorder(pathway, NES), y = NES, fill = NES)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Top Enriched Hallmark Pathways",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)",
    fill = "NES"
  ) +
  theme_minimal(base_size = 14, base_family = "Times New Roman")

```


```{r}
# Plot top pathways NES 
ggplot(topPathways_filtered, aes(x = reorder(pathway, NES), y = NES, fill = NES)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Top Enriched Hallmark Pathways",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)",
    fill = "NES"
  ) +
theme_minimal(base_size = 20, base_family = "Times New Roman") +
theme(
  plot.title   = element_text(hjust = 0.5, size = 16, face = "bold"),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 12),
  axis.title.x = element_text(size = 13),
  axis.title.y = element_text(size = 13),
  legend.title = element_text(size = 12),
  legend.text  = element_text(size = 11),
  panel.grid.major = element_blank(),  
  panel.grid.minor = element_blank()   
)
```

```{r}
topPathways_filtered$pathway_label <- topPathways_filtered$pathway %>%
  str_replace_all("_", " ") %>% 
  str_to_title()                

ggplot(topPathways_filtered, aes(x = reorder(pathway_label, NES), y = NES, fill = NES)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Top Enriched Hallmark Pathways",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)",
    fill = "NES"
  ) +
theme_minimal(base_size = 20, base_family = "Times New Roman") +
theme(
  plot.title   = element_text(hjust = 0.5, size = 16, face = "bold"),
  axis.text.x  = element_text(size = 20),
  axis.text.y  = element_text(size = 12),
  axis.title.x = element_text(size = 13),
  axis.title.y = element_text(size = 13),
  legend.title = element_text(size = 12),
  legend.text  = element_text(size = 11),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
)

```


```{r}
library(stringr)

topPathways_filtered$pathway_label <- topPathways_filtered$pathway %>%
  str_replace_all("_", " ") %>%           # replace underscores with spaces
  str_to_lower() %>%                      # lowercase everything
  str_replace("^hallmark", "Hallmark")    # capitalize "Hallmark" at start

ggplot(topPathways_filtered, aes(x = reorder(pathway_label, NES), y = NES, fill = NES)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Enriched GSEA Hallmark Pathways",
    x = "",
    y = "Normalized Enrichment Score (NES)",
    fill = "NES"
  ) +
theme_minimal(base_size = 20, base_family = "Times New Roman") +
theme(
  plot.title   = element_text(hjust = 0.5, size = 22, face = "bold", color = "black"),
  axis.text.x  = element_text(size = 15, color = "black"),
  axis.text.y  = element_text(size = 15, color = "black"),
  axis.title.x = element_text(size = 13, color = "black", face = "bold"),  
  axis.title.y = element_text(size = 13, color = "black", face = "bold"),  
  legend.title = element_text(size = 15, color = "black"),
  legend.text  = element_text(size = 12, color = "black"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.line = element_line(color = "black", size = 0.5)
)


```



# 10. Extract hallmark genes

```{r}
# fgseaResult already has a column "leadingEdge" that contains gene symbols contributing to the NES
fgsea_gene_list <- fgseaRes[, c("pathway", "NES", "leadingEdge")]

# check if i can access this
fgsea_gene_list$leadingEdge[[1]]
```


```{r}
# all pathways
fgsea_gene_list[[1]]
```


```{r}
# hallmark hypoxia
fgsea_gene_list$leadingEdge[[22]]
```



```{r}
library(EnhancedVolcano)

hallmark_hypoxia_genes <- c(
  "HMOX1", "ADM", "STC1", "ENO2", "STC2", "DDIT4", "LDHC", "CDKN1A", "MT2A", "NDRG1",
  "BNIP3L", "ISG20", "SLC2A3", "PGF", "TMEM45A", "CA12", "PDK1", "BHLHE40", "P4HA1",
  "ALDOC", "GBE1", "MAFF", "KDM3A", "SLC6A6", "EFNA3", "GCNT2", "P4HA2", "KLHL24",
  "ATF3", "BCAN", "ALDOA", "VEGFA", "PLIN2", "PIM1", "HK2", "ANKZF1", "TPST2", "PPP1R3C",
  "GLRX", "ERRFI1", "CCNG2", "ZNF292", "PPP1R15A", "GPI", "CDKN1C", "TES", "PYGM",
  "F3", "NFIL3", "NEDD4L", "ERO1A", "ANXA2", "ATP7A", "PNRC1", "SLC2A1", "NDST1",
  "LDHA", "KIF5A", "SERPINE1", "JMJD6", "IDS", "FOXO3", "JUN", "IER3", "GAPDH",
  "HEXA", "PPARGC1A", "PFKP", "DPYSL4", "EXT1", "AMPD3", "CP", "PLAUR", "PGM1",
  "AKAP12", "CAVIN1", "SDC4", "TIPARP", "PGK1"
)


# Volcano plot labeling only pathway genes
EnhancedVolcano(resLFC,
                lab = resLFC$gene_symbol,
                x = 'log2FoldChange',
                y = 'padj',
                selectLab = hallmark_hypoxia_genes, 
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 3.0,
                labSize = 4.0,
                colAlpha = 0.7,
                legendLabels=c('NS','Log2FC','padj','padj & Log2FC'),
                legendPosition = 'right',
                drawConnectors = TRUE,
                widthConnectors = 0.5,
                colConnectors = 'grey30',
                title = 'Treated vs Control',
                subtitle = 'Volcano plot of DE genes',
                xlim = c(-10, 10))

```



```{r}
## Hypoxia genes - improving the visualisation

# Extract hypoxia pathway genes
hypoxia_genes <- unique(unlist(pathways["HALLMARK_HYPOXIA"]))

# Filter DESeq results for hypoxia genes with |log2FC| ≥ 1.5
highlight_genes <- subset(
  resLFC,
  gene_symbol %in% hypoxia_genes & abs(log2FoldChange) >= 1.5
)
```


```{r}
ggplot(resLFC, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(color = "grey", alpha = 0.7) +

  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "darkgreen", size = 3) +
  ggrepel::geom_text_repel(
    data = highlight_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = gene_symbol),
    color = "darkgreen",
    size = 5,                  
    fontface = "bold",
    family = "Times New Roman",
    segment.color = "darkgreen",
    segment.size = 0.5,
    box.padding = 0.5,
    point.padding = 0.5,
    max.overlaps = Inf
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "black") +
  theme_minimal(base_size = 16, base_family = "Times New Roman") +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title   = element_text(hjust = 0.5, size = 25, face = "bold", color = "black"),
    axis.text.x  = element_text(size = 20, color = "black"),
    axis.text.y  = element_text(size = 20, color = "black"),
    axis.title.x = element_text(size = 22, color = "black", face = "bold"),
    axis.title.y = element_text(size = 22, color = "black", face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.5),
    legend.position = "none"
  ) +
  labs(
    title = "Hypoxia Hallmark Pathway Genes",
    x = "log2(Fold Change)",
    y = "-log10(adjusted p-value)"
  )
```




```{r}
## Mitochondria / Metabolism genes

# Define metabolism-related pathways
metabolism_paths <- c(
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
  "HALLMARK_GLYCOLYSIS",
  "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY"
)

# Extract member genes
metabolism_genes <- unique(unlist(pathways[metabolism_paths]))

# Filter DESeq results for metabolism genes with |log2FC| ≥ 1.5
highlight_genes <- subset(
  resLFC,
  gene_symbol %in% metabolism_genes & abs(log2FoldChange) >= 1.5
)
```


```{r}
ggplot(resLFC, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(color = "grey", alpha = 0.7) +
  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "darkblue", size = 3) +
  ggrepel::geom_text_repel(
    data = highlight_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = gene_symbol),
    color = "darkblue",
    size = 6,                 
    fontface = "bold",
    family = "Times New Roman",
    segment.color = NA,
    box.padding = 0.5,
    point.padding = 0.5,
    max.overlaps = Inf
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "black") +
  theme_minimal(base_size = 16, base_family = "Times New Roman") +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title   = element_text(hjust = 0.5, size = 25, face = "bold", color = "black"),
    axis.text.x  = element_text(size = 20, color = "black"),
    axis.text.y  = element_text(size = 20, color = "black"),
    axis.title.x = element_text(size = 22, color = "black", face = "bold"),
    axis.title.y = element_text(size = 22, color = "black", face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.5),
    legend.position = "none"
  ) +
  labs(
    title = "Mitochondrial and Metabolism Hallmark Pathway Genes",
    x = "log2(Fold Change)",
    y = "-log10(adjusted p-value)"
  )
```

```{r}
## DNA Repair genes

# Define DNA repair pathway
dna_repair_path <- c("HALLMARK_DNA_REPAIR", "HALLMARK_P53_PATHWAY") 

# Extract member genes
dna_repair_genes <- unique(unlist(pathways[dna_repair_path]))

# Filter DESeq results for DNA repair genes with |log2FC| ≥ 1.5
highlight_genes <- subset(
  resLFC,
  gene_symbol %in% dna_repair_genes & abs(log2FoldChange) >= 1.5
)
```

```{r}
library(ggplot2)
library(ggrepel)

ggplot(resLFC, aes(x = log2FoldChange, y = -log10(padj))) +
  # All genes grey
  geom_point(color = "grey", alpha = 0.7) +
  # Highlight DNA repair genes in purple
  geom_point(data = highlight_genes,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "purple", size = 3) +
  # Label highlighted DNA repair genes
  ggrepel::geom_text_repel(
    data = highlight_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = gene_symbol),
    color = "purple",
    size = 6,                 # increased label size
    fontface = "bold",
    family = "Times New Roman",
    segment.color = NA,
    box.padding = 0.5,
    point.padding = 0.5,
    max.overlaps = Inf
  ) +
  # Horizontal significance cutoff line
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  # Vertical fold-change cutoff lines at ±1.5
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "black") +
  theme_minimal(base_size = 16, base_family = "Times New Roman") +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title   = element_text(hjust = 0.5, size = 28, face = "bold", color = "black"),
    axis.text.x  = element_text(size = 20, color = "black"),  # bigger x-axis labels
    axis.text.y  = element_text(size = 20, color = "black"),  # bigger y-axis labels
    axis.title.x = element_text(size = 22, color = "black", face = "bold"),  # bigger x-axis title
    axis.title.y = element_text(size = 22, color = "black", face = "bold"),  # bigger y-axis title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.5),
    legend.position = "none"
  ) +
  labs(
    title = "DNA Repair and p53 Hallmark Pathway Genes",
    x = "log2(Fold Change)",
    y = "-log10(adjusted p-value)"
  )

```



## Innate immune genes


```{r}
# Define innate immunity / inflammation pathways
innate_immunity_paths <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_IL2_STAT5_SIGNALING",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB"
)

# Loop through each pathway to create individual volcano plots
for (pathway in innate_immunity_paths) {
  
  pathway_genes <- unique(unlist(pathways[pathway]))
  
  highlight_genes <- subset(
    resLFC,
    gene_symbol %in% pathway_genes & abs(log2FoldChange) >= 1.5
  )
  
  # Create volcano plot
  p <- ggplot(resLFC, aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(color = "grey", alpha = 0.7) +
    geom_point(data = highlight_genes,
               aes(x = log2FoldChange, y = -log10(padj)),
               color = "darkblue", size = 3) +
    # Label highlighted genes
    ggrepel::geom_text_repel(
      data = highlight_genes,
      aes(x = log2FoldChange, y = -log10(padj), label = gene_symbol),
      color = "darkblue",
      size = 5,
      fontface = "bold",
      family = "Times New Roman",
      segment.color = NA,
      box.padding = 0.5,
      point.padding = 0.5,
      max.overlaps = Inf
    ) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
    theme_minimal(base_size = 14, base_family = "Times New Roman") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
      axis.title = element_text(size = 16),
      axis.text = element_text(size = 14),
      panel.grid = element_blank()
    ) +
    labs(
      title = paste0(pathway, " pathway genes (|log2FC| ≥ 1.5)"),
      x = "log2(Fold Change)",
      y = "-log10(adjusted p-value)"
    )
  
  print(p)
}

```



# Save results

```{r}
library(writexl)

# Save all DEGs with gene symbols
write_xlsx(as.data.frame(resLFC), path = "/path_to_output/DEGs_results_Control_vs_Treat2.xlsx")

# Save only significant DEGs (padj < 0.05)
sig_res <- resLFC[resLFC$padj < 0.05 & !is.na(resLFC$padj), ]
write_xlsx(as.data.frame(sig_res), path = "/path_to_output/DEGs_results_Control_vs_Treat2_significant.xlsx")
```



